### 2024-05-19 fantasy-football.qmd
20240519 - description of football

This chapter provides a brief primer on football and fantasy football.
If you are already familiar with fantasy football, feel free to skip this chapter.

@@

@@

A "flex position" is a flexible position that can involve a player from various positions: e.g., a RB or WR or TE.

@@

| Statistical category                              | Points               |
|---------------------------------------------------|----------------------|
| Rushing or receiving TD                           | 6                    |
| Returning a kick or punt for a TD                 | 6                    |
| Returning or recovering a fumble for a TD         | 6                    |
| Passing TD                                        | 4                    |
| Passing INT                                       | −2                   |
| Fumble lost                                       | −2                   |
| Rushing, passing, or receiving 2 point conversion | 2                    |
| Rushing or receiving yards                        | 1 point per 10 yards |
| Passing yards                                     | 1 point per 25 yards |

@@

| Statistical category                              | Points |
|---------------------------------------------------|--------|
| FG made: 50+ yards                                | 5      |
| FG made: 40–49 yards                              | 4      |
| FG made: 39 yards or less                         | 3      |
| Rushing, passing, or receiving 2 point conversion | 2      |
| Point after attempt made                          | 1      |
| Point after attempt missed                        | −1     |
| Missed FG: 0–39 yards                             | −2     |
| Missed FG: 40–49 yards                            | −1     |

@@

| Statistical category  | Points |
|-----------------------|--------|
| Tackle solo           | 1      |
| Tackle assist         | 0.5    |
| Tackle for loss       | 1      |
| Sack                  | 2      |
| Interception          | 4      |
| Fumble force          | 2      |
| Fumble recovery       | 2      |
| TD                    | 6      |
| Safety                | 2      |
| Pass defended         | 1      |
| Block kick            | 2      |
| Extra point returned  | 2      |

@@

- "TD" = touchdown
- "INT" = interception
- "yds" = yards
- "ATT" = attempts
- "2 pt conversion" = two-point conversion
- "FG" = field goal
- "PAT" = point after touchdown (i.e., extra point/point after attempt)

>>>>

This chapter provides a brief primer on (American) football and fantasy football.
If you are already familiar with fantasy football, feel free to skip this chapter.

@@

The goal in football is for a team to score more points than their opponent.

### The Roster {#sec-footballRoster}

Each team has 11 players on the field at a time.
The particular players who are on the field will depend on the situation, but usually includes one of the three subsets of players:

1. Offense
1. Defense
1. Special Teams

#### Offense {#sec-footballRosterOffense}

The offense is on the field when the team has the ball.

Players on offense include:

- quarterback (QB)
- running back (RB), also called halfback or tailback
- fullback (FB)
- wide receiver (WR)
- tight end (TE)
- offensive line (OL)
    - center (C)
    - offensive guard (OG)
    - offensive tackle (OT)

Quarterbacks are the most important player on the offense.
They help lead the team down the field.
Quarterbacks receive the ball from the Center at the beginning of the play, and they can either hand the ball off (typically to a running back or fullback), pass the ball (typically to a wide receiver or tight end), or run the ball.

Running backs take a hand-off from the quarterback to execute a running play.
They may also catch short passes from the quarterback or help protect (i.e., block for) the quarterback from the defensive players who are trying to tackle the quarterback.

Wide receivers catch passes from the quarterback to execute a passing play.
On running plays, they provide protection for the player running the ball (e.g., the running back) so they can get as far as possible without being tackled.

Tight ends block for running and passing plays, and they catch passes from the quarterback.

The offensive line blocks for running and passing plays.
On passing plays, they provide protection for the quarterback so the quarterback has time to pass the ball without being tackled.
On running plays, they provide protection for the player running the ball (e.g., the running back) so they can get as far as possible without being tackled.

#### Defense {#sec-footballRosterDefense}

The defense is on the field when the team does not have the ball (i.e., when the opposing team has the ball).

Players on defense include:

- defensive line (DL)
    - defensive end (DE)
    - defensive tackle (DT)
- linebacker (LB)
    - middle linebacker (MLB)
    - outside linebacker (OLB)
- defensive back (DB)
    - cornerback (CB)
    - safety (S)
        - free safety (FS)
        - strong safety (SS)

#### Special Teams {#sec-footballRosterSpecialTeams}

The special teams involves specialist players who are on the field during all kicking plays including kickoffs, field goals, and punts.

Players on special teams include:

- kicker (K)
- punter (P)
- holder
- long snapper
- punt returner
- kick returner
- and other players intended to block for or to tackle the ball carrier

### The Field {#sec-footballField}

The football field is rectangular and is 120 yards long and 53 1/3 yards wide.
At each end of the 120-yard field is a team's end zone.
Each end zone is 10 yards long.
Thus, the distance from one end zone to the other end zone is 100 yards.
Behind each end zone is a field goal post.

### The Gameplay {#sec-footballGameplay}

At the beginning of the game, there is a coin flip to determine which teams receives the ball first and which team takes which side of the field.
During the kickoff, the kicking team kicks the ball to the receiving team, who has the option to return the kick.
The offense starts their possession at the 25 yard line—if there is no return (i.e., a touchback)—or wherever the kick returner is tackled or goes out of bounds.

The team with the ball (i.e., the offense) has four opportunities ("downs") to advance the ball 10 yards.
A team can advance the ball either by running it or passing and catching it.
At the end of a rushing play, the ball advances to wherever the ball carrier is tackled or goes out of bounds (i.e., wherever the player is "down").
At the end of a passing play, if the thrown ball is caught (i.e., a completed pass), the ball advances to wherever the ball carrier is tackled or goes out of bounds.
If the thrown ball is not caught (i.e., an incomplete pass), the ball does not advance.
Wherever the ball is advanced to dictates where the next play begins.
The yard position on the field where the next play takes place from is known as the line of scrimmage.
Neither team can cross the line the line of scrimmage until the next play begins.
To begin the play, the ball is placed on the line of scrimmage and the Center gives (or "snaps") the ball to the Quarterback.

If the team advances the ball 10 or more yards within four downs, the team is awarded a new set of downs and is given four more downs to advance the ball 10 more yards.
If the team advances the ball all the way to the other team's end zone, they score a touchdown.
If the team fails to advance the ball 10 or more yards within four downs, the team loses the ball, and the other team takes possession at that spot on the field.
There are risks of giving the other team the ball with a short distance to score.
Thus, on fourth down, instead of trying to advance the ball for a first down, a team may choose to kick a field goal—to get points—or to punt.

A field goal involves a kicker kicking the ball with an intent to kick the ball through the field goal posts ("uprights").
To achieve points, the field goal must be above the cross bar and be in between the uprights (extended vertically) when it crosses the cross bar.

Punting involves a punter kicking the ball to the other team with an intent to give their opponent worse field position, thus making it harder for the other team to score.
The punting team tries to pin the opponent as close as possible to the opponent's end zone (i.e., as far as possible from the own team's end zone), so they have a longer distance to go to score a touchdown.

There are multiple ways that ball possession can switch from the offense to the other team.
After scoring a touchdown, field goal, or safety, there is a kickoff, in which the scoring team kicks the ball to the opponent.
Another way that the ball swithces possession to the other team is if the team commits a turnover.
The defense can force a turnover by an interception, fumble recovery, or turnover on downs.
A turnover due to an interception occurs when the defense catches the the quarterback's pass.
A turnover due to a fumble recovery occurs when an offensive player, who had possession of the ball, loses the ball before being down or scoring a touchdown and the ball is recovered by the opponent.
A turnover on downs occurs when the team attempts to achieve the remainder of the needed 10 yards to go on fourth down but fails.

Other football-related situations include tackles for loss and sacks.
A tackle for loss occurs when a (non-quarterback) is tackled behind the line of scrimmage.
A sack occurs when a quarterback is tackled behind the line of scrimmage.
A pass defended occurs when a defensive back knocks down the ball in the air so that the indended receiver cannot catch the ball.

### The Scoring {#sec-footballScoring}

The goal of the team with the ball (i.e., the offense) is to score points.
It can do this by either advancing the ball into the other team's end zone (6 points) or by kicking a field goal (3 points).
Advancing the ball in the other team's end zone is called a touchdown.
After a touchdown, the offense chooses to attempt either a point-after-touchdown (PAT) or a two-point conversion.
A PAT is a short kick attempt from the 15-yard line that, if it goes through the goal posts ("uprights"), is worth 1 point.
A two-point conversion is a single-scoring opportunity from the 3-yard line (i.e., 3 yards away from the end zone).
If the offense scores (i.e., advances the ball into the end zone) from the 3-yard line, the team is awarded 2 points.

A team can kick a field goal from any distance as long as the kick goes through the goal posts.
The current record for the longest field goal is 66 yards (by Justin Tucker in 2021).

A safety occurs when the offense is tackled with the ball in their own end zone.
When a safety occurs, the opposing team (i.e., defense) is awarded two points.

### Glossary of Terms {#sec-footballGlossary}

- running play ("run") or rushing play (or "rush")
- passing play (or "pass")
- passing attempt
- rushing attempt
- passing completion
- passing incompletion
- passing yards
- rushing yards
- receving yards
- reception
- touchdown
- passing touchdown
- rushing touchdown
- receiving touchdown
- two-point conversion
- block
- kickoff
- field goal
- point after touchdown (PAT)
- extra point returned
- punt
- fumble lost
- fumble forced
- fumble recovery
- interception
- tackle
- tackle solo
- tackle assist
- tackle for loss
- sack
- pass defended
- safety

@@

A "flex position" is a flexible position that can involve a player from various positions: e.g., a running back, wide receiver, or tight end.
In fantasy football, fullbacks are classified as running backs because they both run the ball.

@@

| Statistical category                              | Points               |
|---------------------------------------------------|----------------------|
| Rushing or receiving TD                           | 6                    |
| Returning a kick or punt for a TD                 | 6                    |
| Returning or recovering a fumble for a TD         | 6                    |
| Passing TD                                        | 4                    |
| Passing INT                                       | −2                   |
| Fumble lost                                       | −2                   |
| Rushing, passing, or receiving 2-point conversion | 2                    |
| Rushing or receiving yards                        | 1 point per 10 yards |
| Passing yards                                     | 1 point per 25 yards |

@@

| Statistical category                              | Points |
|---------------------------------------------------|--------|
| FG made: 50+ yards                                | 5      |
| FG made: 40–49 yards                              | 4      |
| FG made: 39 yards or less                         | 3      |
| Rushing, passing, or receiving 2-point conversion | 2      |
| Point after touchdown attempt made                | 1      |
| Point after touchdown attempt missed              | −1     |
| Missed FG: 0–39 yards                             | −2     |
| Missed FG: 40–49 yards                            | −1     |

@@

| Statistical category  | Points |
|-----------------------|--------|
| Tackle solo           | 1      |
| Tackle assist         | 0.5    |
| Tackle for loss       | 1      |
| Sack                  | 2      |
| Interception          | 4      |
| Fumble forced         | 2      |
| Fumble recovery       | 2      |
| TD                    | 6      |
| Safety                | 2      |
| Pass defended         | 1      |
| Blocked kick          | 2      |
| Extra point returned  | 2      |

@@

- "TD" = touchdown
- "INT" = interception
- "yds" = yards
- "ATT" = attempts
- "2-pt conversion" = two-point conversion
- "FG" = field goal
- "PAT" = point after touchdown (i.e., extra point/point after attempt)
